---
title: "Part 5"
subtitle: "R Programming"
author: "Jessica Minnier"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [css/xaringan-themer.css, css/my-theme.css]
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
mono_light(
  base_color =  "#3A6185", ## OHSU Marquam
  code_highlight_color = "#cbdded",
  link_color = "#38BDDE",
  header_font_google = google_font("Josefin Sans"),
  text_font_google   = google_font("Montserrat", "300", "300i","400i","700"),
  code_font_google   = NULL,
  text_font_size = "24px",
  code_font_size = "20px",
  header_h1_font_size = "45px",
  header_h2_font_size = "40px",
  header_h3_font_size = "35px",
  padding = "0em 2em 1em 2em",
  outfile = "css/xaringan-themer.css"
)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "logo", "tachyons"))
xaringanExtra::use_panelset()
xaringanExtra::use_tachyons()
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```

class: center, middle

# Saving objects

---

# Penguins data

```{r, message=FALSE}
library(tidyverse)
library(janitor)
library(palmerpenguins)
data(penguins)
glimpse(penguins)
```

---

# When we filter/clean/mutate the data

- If you're changing the data and you want to use that data later, save your work
- Save as new object, *or*,
- Save as the same object if you want to overwrite that data frame within R


```{r}
# save as new data frame penguins_f
penguins_f <- penguins %>%
  filter(sex=="female") %>%
  mutate(bill_ratio = bill_length_mm/bill_depth_mm)

# change the ordering of penguins_f and re-save it
penguins_f <- penguins_f %>% arrange(bill_ratio)

# print out that data frame, but don't save it
penguins_f %>% select(sex, contains("bill")) 
```

---

# Summarizing data (i.e. tabyl, summarize)

- `tabyl()` and `summarize()` create new tibbles/data.frames that have summary info
- You can save it if you want, but not usually necessary unless you want to use it later!
- From `?tabyl`, Value = "Returns a data.frame with frequencies and percentages of the tabulated variable(s). A 3-way tabulation returns a list of data.frames."

.pull-left-40[
```{r}
penguins_f %>% tabyl(species)
```
]
.pull-right-60[
```{r}
# save output as a tibble
penguins_f_table <- penguins_f %>% tabyl(species)

# we can see this is a tibble!
glimpse(penguins_f_table) 

class(penguins_f_table)
```
]

---

# Using summaries later


```{r}
penguins_f_table %>% arrange(n)
```

Nice printing:

```{r}
penguins_f_table %>% gt::gt()
```

---

# Using summaries later

You can also save your summaries to excel or csv files:

```{r}
# one long pipe
penguins_f_table %>% 
  arrange(n) %>%
  write_csv(file = "table_of_species_in_females.csv")

# or save as tibble object and then write tibble to csv file
penguins_f_table <- penguins_f_table %>% arrange(n)

write_csv(penguins_f_table, file = "table_of_species_in_females.csv")
```

---

# "Printing" or "viewing"

Some things you *__don't__* usually want to save

- `glimpse()` prints a snapshot of your data, not useful to save
- `head()` or `tail()`, there's almost no reason to save just the first few lines of your data, except for educational purposes!
- Similarly, you can pipe on `nrow()` or `dim()` or `colnames()` or `class()` and so on, these are pulling out meta-data/information from your data; you can save this as an object but make sure you know what the output is! (probably a vector of information)
- `skim()` prints a visual summary of your data, you can save it if you want to do fancy things with it
- `view()`/`View()` pops up a window that shows your data, don't save this and don't keep this code in your Rmd

---

### glimpse

```{r}
# when you save this, it prints it, and then you could print it again
glimpse_out <- glimpse(penguins_f)
```

Print it now (but really, don't do this, there's no point):

```{r}
glimpse_out
```

---

### skim

from `?skim`: Value = "A skim_df object, which also inherits the class(es) of the input data. In many ways, the object behaves like a `tibble::tibble()`."

```{r}
# when you save this, it doesn't print it, but you can print it later
skim_out <- skimr::skim(penguins_f)
```


Now you can print it (this gets cut off on the slide)

```{r}
skim_out
```


---

You can `glimpse()` the skim output because it's deep down a tibble!

```{r}
class(skim_out)
glimpse(skim_out)
```

---

See `?skim` for examples on what you can do with `skim()` output, as well as the [skimr vignette](https://cran.r-project.org/web/packages/skimr/vignettes/skimr.html)

```{r}
skim_out %>% filter(n_missing > 0)
skim_out %>% 
  filter(skim_type == "numeric") %>%
  select(skim_variable, numeric.mean)
```


---

### view

- Use `View` or `view()` interactively, don't include this code in your Rmd, and don't save it as an object!

```{r, eval = FALSE}
View(penguins_f) # this will pop up every time you knit, not necessary
pview <- penguins_f %>% View() # this is not helpful
```

```{r include = FALSE}
# CHEATING =)
pview = NULL
```


pview is `NULL` because `View()` produces NO output (no value)

```{r}
pview 
```

---

# Saving ggplot

- Save a ggplot *as an object* if you want to "call it" later to add more layers, or if you want to save it *as an image file* (i.e .png, .pdf, .tif, etc) on your computer
- `ggsave()` by default saves the last plot printed to a file, but to be extra careful you can use the object name

```{r, fig.width = 10, fig.height = 3}
# save as object p_boxplot; will not print out plot
p_boxplot <- ggplot(penguins, 
                    aes(x = species,
                        y = bill_length_mm,
                        fill = sex)) +
  geom_boxplot()

# it's a ggplot object
class(p_boxplot)
# print the plot in output
p_boxplot
```

---

Now we can add additional layers, and/or save it as a png file!

```{r, fig.width = 10, fig.height = 3}
# save as png (default location is same folder where Rmd is)
ggsave(p_boxplot, file = "p_boxplot.png", height = 6, width = 5)
# add layers
p_boxplot + theme_minimal() + labs(title = "Add layers to my boxplot!", x = "")
```

---

Want to save that plot?

```{r, fig.width = 10, fig.height = 3}
p_boxplot <- p_boxplot + theme_minimal() + labs(title = "Add layers to my boxplot!", x = "")
ggsave(p_boxplot, file = "p_boxplot_minimal.png", height = 6, width = 5)

p_boxplot
```

---

# Objects

- All of this has to do with classes and objects in R, which is an object-oriented programming language.
- `class(objectname)` tells us what kind of structure it is and how we can use it
- Many functions output a unique kind of object

```{r}
tmpfit <- lm(bill_length_mm ~ species, data = penguins)
class(tmpfit)
# see Value section of ?lm
```

- If you want some bedtime reading on this, start with the [Advanced R textbook's "OO field guide"](http://adv-r.had.co.nz/OO-essentials.html)
