---
title: "Class 2 Intro to Working with Data: for loops, Rstudio proj, tidy data, ggplot intro"
subtitle: "R Programming"
author: "Jessica Minnier"
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: xaringan-themer.css
    nature:
      ratio: '16:9'
      highlightStyle: github
      highlightLines: true
      countIncrementalSlides: false
---

```{r xaringan-themer, include = FALSE}
library(xaringanthemer)
style_mono_accent(
  base_color = "midnightblue",
  header_font_google = google_font("Josefin Sans"),
  link_color = "steelblue",
  text_font_size = "26px",
  footnote_font_size = "22px"
)
```

```{r xaringanExtra, echo=FALSE}
xaringanExtra::use_xaringan_extra(c("tile_view", "logo", "tachyons"))
xaringanExtra::use_panelset()
xaringanExtra::use_tachyons()
```

```{r setup, include=FALSE}
options(htmltools.dir.version = FALSE)
```


# Class 2

- Muddiest/clearest points
- Some details
    * Youtube video (on website) vs. zoom video link (in slack)
- Go back to for loops from Class 1 Rmd
    * another example
    * `print()`-ing what's going on
- Function of the week
    * Open new project in existing folder
    * Sign up sheet and details
- Class 2 Rproj/Rmd
    * Rstudio projects and file paths
    * Importing data
    * Tidy data
    * Ggplot *__intro__* (advanced stuff later)
- Assignment 2 due date depends on how far we get today
    

---


class: center, middle

# For Loops

---

<img src = "image/week2/for_loop1.jpg" width = 800>

---

<img src = "image/week2/for_loop2.jpg" width = 800>

---

<img src = "image/week2/for_loop3.jpg" width = 800>

---

<img src = "image/week2/for_loop.jpg" width = 800>

[Image from Allison Horst](https://github.com/allisonhorst/stats-illustrations)

---
class: center, middle

# Paths and Projects

---

<img src = "image/week2/rstudio_proj_pic.png" width = 900>

---
class: center, middle

# Tidy Data

---

<img src = "image/week2/tidydata_1.jpg" width = 800>

[Image from Allison Horst](https://github.com/allisonhorst/stats-illustrations)
---

<img src = "image/week2/tidydata_2.jpg" width = 800>

[Image from Allison Horst](https://github.com/allisonhorst/stats-illustrations)

---
class: center, middle

# ggplot2 slides
---

<img src = "image/week2/ggplot2_exploratory.png" width = 600>

[Image from Allison Horst](https://github.com/allisonhorst/stats-illustrations)

---

<img src = "image/week2/khealy_ggplot1_part1.jpg" width = 700>

Image from Kieran Healy