---
title: "Part 2: Loading Data, data.frames, and ggplot2"
linktitle: "Part 2: Loading Data, data.frames, and ggplot2"
date: "2023-01-18"
class_date: "2023-01-18"
output:
  blogdown::html_page:
    toc: true
menu:
  class:
    parent: Class sessions
    weight: 2
type: docs
weight: 2
# pdf: 
# thumb: 
editor_options: 
  chunk_output_type: console
---

## R Project files

Please download the `part2` folder from [this dropbox folder link](https://bit.ly/bsta504_dropbox_2023) Be sure to unzip if necessary. In advance of class, please open the `part2` Rstudio project (double click on the `.rproj` file), open `part2.Rmd` and `knit` (click the Knit button at the top of the file) this file. This will install packages that you need for the Rmd to run.

## Readings

Required and suggested class readings can be found on the Readings tab by class. These readings may be done anytime before or after class, but they will supplement your understanding of the class materials and help make homework and project work easier.


## This year's class video

See Slack for the zoom recording link


## Last Year's Class Video

<iframe width="560" height="315" src="https://www.youtube.com/embed/0-_luKuYo_E" frameborder="0" allow="accelerometer; autoplay; clipboard-write; encrypted-media; gyroscope; picture-in-picture" allowfullscreen></iframe>

View last year's class and materials [here](https://sph-r-programming-2022.netlify.app/class/02-class/).

# Post-Class

Please fill out the following survey and we will discuss the results during the next lecture. All responses will be anonymous.

- Clearest Point: What was the most clear part of the lecture?
- Muddiest Point: What was the most unclear part of the lecture to you?
- Anything Else: Is there something you'd like me to know?

<https://bit.ly/bsta504_postclass_survey>


# Muddiest Points

> the benefits of tibble vs data frame and when to use which?

In this class we will always use tibble. Just remember that an object can be multiple types. A tibble is a data frame, but not vice versa. A tibble is really a data frame with "perks". See this explanation from the [tibble 1.0 package release](https://posit.co/blog/tibble-1-0-0/#:~:text=Tibbles%20vs%20data%20frames,to%20work%20with%20large%20data.)

> There are two main differences in the usage of a data frame vs a tibble: printing, and subsetting.

> Tibbles have a refined print method that shows only the first 10 rows, and all the columns that fit on screen. This makes it much easier to work with large data. In addition to its name, each column reports its type, a nice feature borrowed from `str()`:

```{r}
library(tidyverse)

class(mtcars)
mtcars

mtcars_tib <- as_tibble(mtcars)
class(mtcars_tib)
mtcars_tib
```

Another interesting difference is that tibbles don't have row names, but a lot of built in data.frames in R do. But rownames are hard to get out. So, when you make a tibble of a data.frame you can tell the function to use the rownames as a column:

```{r}
mtcars_tib <- as_tibble(mtcars, rownames = "car")
mtcars_tib
```


> Tibbles also clearly delineate [ and [[: [ always returns another tibble, [[ always returns a vector. No more drop = FALSE!

If we ask for the first column using the `[]` notation, we receive a numeric vector from a data frame, and a tibble/data.frame from the tibble.

We have not learned the `[[]]` yet because we have not talked about lists in R, but we will soon. The code below returns the first column as a vector for both a data frame and a tibble.

```{r}
mtcars[,1]
mtcars_tib[,1]
mtcars[[1]]
mtcars_tib[[1]]

class(mtcars[,1])
class(mtcars_tib[,1])
class(mtcars_tib[[1]])
```

As I was mentioning in class, there are some (older) functions that don't like tibbles, but all you need to do is just make its primary class a data.frame as such:

> A handful of functions are don’t work with tibbles because they expect df[, 1] to return a vector, not a data frame. If you encounter one of these functions, use as.data.frame() to turn a tibble back to a data frame:

```{r}
mtcars_df <- as.data.frame(mtcars_tib)
class(mtcars_df)
```

Back to muddy quotes:

> path files and knowing if you're in a project or just an RMD

> R markdown vs R projects

I hope to spend more time talking about this in class 4.

> ggplot stuff was the most muddy, but I also haven't done a lot of ggplot stuff before

Yes this was definitely expected for a brief intro, ggplot takes a while to get the hang of! We will use ggplot every class now, so we will go through it in bite sized pieces.

> Using na="NA" to pull in data and how to know that it's needed.

I will show more examples of this. Rule number one of importing data in any software is to *look* at your data, and figure out if what you see in the software is what you expect. Always look at your data! The `read_excel(filename, na="NA")` is a strange case that isn't actually very common to code data as "NA" directly, but I wanted to show you how it looks different when it does happen. Usually, missing data is just a blank space, which is automatically read in as the special `NA` data type in R.

```{r}
# If you did not include `na=NA` it would have been read in like this
df1 <- tibble(a = c("NA","C","D"), b= 1:3,  c = c(1,3,"NA"))
# If you did include `na = NA` it would have been read in like this
df2 <- tibble(a = c(NA,"C","D"), b= 1:3, c = c(1,3,NA))

# note the character types of the two DFs, and the way NA is printed
df1

df2

```


> I saw a lot of code with the two colons (“::”) in the middle. It is unclear to me if this is an alternative way to write some commands or if there is a certain context in which it is used.

Good question, what this does is pulls a function from a package, so it works whether you have *loaded* the package (using `library()` or `p_load()`) or not. I mainly use it as a clue to you to where the function is coming from. Otherwise, you may not know you need to load that package to use it! For instance:

```{r, eval = FALSE}
# does not work, haven't loaded the package janitor
mtcars %>% tabyl(disp, cyl)
```


```{r}
# does work
mtcars %>% janitor::tabyl(disp, cyl)
```

```{r}
# also works
library(janitor)
mtcars %>% tabyl(disp, cyl)
```

# Clearest Points

> skim

> loading our excel to R studio

> Loading in the data and selecting the sheets that are most relevant to what we are looking to do was very clear and a nice foundation for future projects. I found that showing different ways of importing the data was helpful.

I'm glad, the import tool in Rstudio is very nice.

> functionality of ggplot

> tidying the data

> Found out what eval=TRUE and eval=FALSE mean!

Great and I'll show that again for anyone who was confused! ("still a little bit confused about the `{r, EVAL}` code")

# Other messages

Some people had trouble getting the `fig.path=` to work in the knitr options. I'm not sure what could be causing that but feel free to ask me during break.

[Here's a good reference](https://yihui.org/knitr/options/) for all the code chunk options, if you want to read about it.


> link to the course website that is in the overview tab in SAKAI links to last years materials.

Oops thank you great catch, fixed!

> Speed is going great. I'm just worried as we progress through the course, it'll be more difficult. Overall, really enjoying this class.

I understand the concern, some things will get more difficult (I'm thinking `across()` in class 4, writing functions, and `purrr`), but we will also circle back to some things that might be familiar or maybe less complicated to start (stats models, making tables). Definitely keep asking questions and I will slow down as needed!

